---
title: "Microbiome_IBR_sharing"
author: "Sarah Vogel"
date: "2025-5-5"
output: html_document
---

#setup

##load libraries
```{r}
suppressPackageStartupMessages({
  library(vegan)
  library(tidyverse)
  library(stats)
  library(phyloseq)
  library(mefa)
  library(qiime2R)
  library(microbiome)
  library(fantaxtic)
  library(car)
  library(ggpubr)
  library(readxl)
  library(psych)
  library(sjPlot)
  library(reshape2)
  library(mosaic)
  library(Maaslin2)
  library(lm.beta)
  library(lavaan)
  library(lcsm)
  library(dplyr)
  library(tidyr)
  library(readr)
  library(DirichletMultinomial)
  library(reshape2)
  library(magrittr)
  library(tibble)
})
```

##load in data files
```{r}
df <- read.csv("metadata_all_bf_sharing.csv")
df$X <- NULL
```


#Creating affect composite
Note - R markdown seems to read Lavaan models strangely when you run them in chunks. The best way to execute these models is to highlight the whole model (i.e. lines 62-64) and run that, then execute the fit and summary commands as normal.  
```{r}
#measurement model
measurement <- '
temp =~ ibr11_T30 + ibr4_T30 + ibr5_T30 
'
fitm <- cfa(measurement,fixed.x=F,data = df)
summary(fitm, fit.measures = TRUE, standardized = TRUE, rsquare=TRUE, estimates = TRUE, ci = TRUE)
```

##extract LV affect for use in MaAslin models
```{r}
affect <- lavPredict(fitm)

df2 <- df[!is.na(df$ibr11_T30)& !is.na(df$ibr4_T30) & !is.na(df$ibr5_T30),]

df2 <- cbind(df2, affect)

write.csv(df2, "metadata_all_LV_sharing.csv")
write_tsv(df2, "metadata_all_LV_sharing.tsv")
```

#Alpha diversity analyses
##shannon
```{r}
solutions <- '
temp ~ shannon_2 + tc_sex + shannon_18 + SES + months_bf_T18 + PH_cesarean_T2  
'
fits <- cfa(solutions, fixed.x=F,data = df2)
summary(fits, fit.measures = TRUE, standardized = TRUE, rsquare=TRUE, estimates = TRUE, ci = TRUE)
```
##chao1
```{r}
solutionc <- '
temp ~ chao1_2 + chao1_18 + tc_sex +SES + months_bf_T18 + PH_cesarean_T2
'
fitc <- cfa(solutionc, fixed.x=F,data = df2)
summary(fitc, fit.measures = TRUE, standardized = TRUE, rsquare=TRUE, estimates = TRUE, ci = TRUE)

```

#Differential abundance models, exploratory
##2 weeks
###species
```{r}
LV30_spec_2wk <- Maaslin2(
  input_data = "2week_species_abundances.tsv",
  input_metadata = "metadata_all_LV.tsv",
  output = "~/DA/2weeks/species",
  min_abundance = 0,
  min_prevalence = 0.5, 
  max_significance = 0.25, 
  analysis_method = "lm", #default
  fixed_effects = c("temp"),
  standardize = TRUE
)
```
###genus
```{r}
LV30_genus_2week <- Maaslin2(
  input_data = "2week_genusabundances_DA.tsv",
  input_metadata = "metadata_all_LV.tsv",
  output = "~/Dropbox/BASELab/DMMProject/Data/Analyses/DA/2weeks/Genus",
  min_abundance = 0,
  min_prevalence = 0.5,
  max_significance = 0.25, 
  analysis_method = "lm", #default
  fixed_effects = c("temp"),
  standardize = TRUE
)
```

##18 months
###species
```{r}
LV30_spec_18mo <- Maaslin2(
  input_data = "18mo_DAready_species.tsv",
  input_metadata = "metadata_all_LV.tsv",
  output = "~/Dropbox/BASELab/DMMProject/Data/Analyses/DA/18mo/species",
  min_abundance = 0,
  min_prevalence = 0.5, 
  max_significance = 0.25, 
  analysis_method = "lm", #default
  fixed_effects = c("temp"),
  standardize = TRUE
)
```

###genus
```{r}
LV30_genus_18mo <- Maaslin2(
  input_data = "18mo_genusabundances_DA.tsv",
  input_metadata = "metadata_all_LV.tsv",
  output = "~/Dropbox/BASELab/DMMProject/Data/Analyses/DA/18mo/Genus",
  min_abundance = 0,
  min_prevalence = 0.5, 
  max_significance = 0.25, 
  analysis_method = "lm", #default
  fixed_effects = c("temp"),
  standardize = TRUE
)
```

#Taxa linear models with controls
##format data for linear models
###2 week
```{r}
#genus, want to grab genera associated at 2 weeks and 18 months for use as controls
genus_features2wk <- read_tsv("2week_genusabundances_DA.tsv")
#subset to only taxa of interest, those identified as significant in MaAslin models

genus2wk_sub <- genus_features2wk[, c("SeqID", "k__Bacteria.p__Firmicutes.c__Negativicutes.o__Veillonellales.f__Veillonellaceae.g__Veillonella", "k__Bacteria.p__Firmicutes.c__Clostridia.o__Eubacteriales.f__Clostridiaceae.g__Clostridium", "k__Bacteria.p__Firmicutes.c__Bacilli.o__Lactobacillales.f__Streptococcaceae.g__Streptococcus", "k__Bacteria.p__Firmicutes.c__Clostridia.o__Eubacteriales.f__Oscillospiraceae.g__Anaerotruncus", "k__Bacteria.p__Firmicutes.c__Clostridia.o__Eubacteriales.f__Oscillospiraceae.g__Ruthenibacterium", "k__Bacteria.p__Firmicutes.c__Bacilli.o__Lactobacillales.f__Lactobacillaceae.g__Lactobacillus", "k__Bacteria.p__Actinobacteria.c__Actinomycetia.o__Bifidobacteriales.f__Bifidobacteriaceae.g__Bifidobacterium", "k__Bacteria.p__Bacteroidetes.c__Bacteroidia.o__Bacteroidales.f__Bacteroidaceae.g__Bacteroides")]

#rename columns to genus name and with _2 to distinguish from 18 month
colnames(genus2wk_sub) <- c("SeqID", "veillonella_2", "clostridium_2", "streptococcus_2", "anaerotruncus_2", "ruthenibacterium_2", "lactobacillus_2", "bifidobacterium_2", "bacteroides_2")

df_gens <- merge(df2, genus2wk_sub, by = "SeqID")

```

###18 month
```{r}
#18 month
genus_18mo <- read_tsv("18mo_genusabundances_DA.tsv")
#subset to only taxa of interest, those identified in the 2 week MaAslin models
genus_18mo_sub <- genus_18mo[, c("SeqID", "k__Bacteria.p__Firmicutes.c__Negativicutes.o__Veillonellales.f__Veillonellaceae.g__Veillonella",
                                 "k__Bacteria.p__Firmicutes.c__Clostridia.o__Clostridiales.f__Clostridiaceae.g__Clostridium", 
                                 "k__Bacteria.p__Firmicutes.c__Bacilli.o__Lactobacillales.f__Streptococcaceae.g__Streptococcus",
                                 "k__Bacteria.p__Firmicutes.c__Clostridia.o__Clostridiales.f__Ruminococcaceae.g__Ruthenibacterium", 
                                 "k__Bacteria.p__Firmicutes.c__Clostridia.o__Clostridiales.f__Ruminococcaceae.g__Anaerotruncus",
                                 "k__Bacteria.p__Firmicutes.c__Bacilli.o__Lactobacillales.f__Lactobacillaceae.g__Lactobacillus",
                                 "k__Bacteria.p__Actinobacteria.c__Actinobacteria.o__Bifidobacteriales.f__Bifidobacteriaceae.g__Bifidobacterium",
                                 "k__Bacteria.p__Bacteroidetes.c__Bacteroidia.o__Bacteroidales.f__Bacteroidaceae.g__Bacteroides")]

colnames(genus_18mo_sub) <- c("SeqID", "veillonella_18", "clostridium_18", "streptococcus_18","ruthenibacterium_18", "anaerotruncus_18", "lactobacillus_18", "bifidobacterium_18", "bacteroides_18")

df_gens <- merge(df_gens, genus_18mo_sub, by = "SeqID", all = TRUE)

#do some log transformations of the abundance variables

df_gens$streptococcus_2_log <- log(df_gens$streptococcus_2 + 0.0001)
describe(df_gens$streptococcus_2_log)#no longer skewed

df_gens$veillonella_2_log <- log(df_gens$veillonella_2 + 0.0001)
describe(df_gens$veillonella_2_log)#no longer skewed

df_gens$clostridium_2_log <- log(df_gens$clostridium_2 + 0.0001)
describe(df_gens$clostridium_2_log)#no longer skewed

df_gens$streptococcus_18_log <- log(df_gens$streptococcus_18 + + 0.0001)
df_gens$veillonella_18_log <- log(df_gens$veillonella_18 + + 0.0001)
df_gens$clostridium_18_log <- log(df_gens$clostridium_18 + + 0.0001)

df_gens$anaerotruncus_2_log <- log(df_gens$anaerotruncus_2 + + 0.0001)
df_gens$ruthenibacterium_2_log <- log(df_gens$ruthenibacterium_2 + + 0.0001)

#bifidobacterium 2 and 18 not skewed

df_gens$lactobacillus_2_log <- log(df_gens$lactobacillus_2 + + 0.0001)
df_gens$bacteroides_2_log <- log(df_gens$bacteroides_2 + + 0.0001)
df_gens$lactobacillus_18_log <- log(df_gens$lactobacillus_18 + + 0.0001)

#bacteroides 18 not skewed

df_gens$anaerotruncus_18_log <- log(df_gens$anaerotruncus_18 + + 0.0001)
df_gens$ruthenibacterium_18_log <- log(df_gens$ruthenibacterium_18 + + 0.0001)

```

##Run models

###bifidobacterium
```{r}
bifido <- '
temp ~ bifidobacterium_2 + bifidobacterium_18 + tc_sex + SES + months_bf_T18+ PH_cesarean_T2
'
fitb <- sem(bifido, fixed.x = F, data = df_gens)
summary(fitb, fit.measures = TRUE, standardized = TRUE, rsquare=TRUE, estimates = TRUE, ci = TRUE)

```

###lactobacillus
```{r}
lacto <- '
temp ~ lactobacillus_2_log + lactobacillus_18_log + tc_sex + SES + months_bf_T18+ PH_cesarean_T2 
'
fitl <- sem(lacto, fixed.x = F, data = df_gens)
summary(fitl, fit.measures = TRUE, standardized = TRUE, rsquare=TRUE, estimates = TRUE, ci = TRUE)

```

###bacteroides
```{r}
bacter <- '
temp ~ bacteroides_2_log + bacteroides_18 + tc_sex + SES + months_bf_T18+ PH_cesarean_T2
'
fitbac <- sem(bacter, fixed.x = F,data = df_gens)
summary(fitbac, fit.measures = TRUE, standardized = TRUE, rsquare=TRUE, estimates = TRUE, ci = TRUE)

```



###veillonella
```{r}
 #controlling for 18 mo, SES, tcsex, 

modelv <- '
temp ~ veillonella_2_log + veillonella_18_log + tc_sex + SES + months_bf_T18+ PH_cesarean_T2
'
fitv <- sem(modelv, fixed.x=F,data = df_gens)
summary(fitv, fit.measures = TRUE, standardized = TRUE, rsquare=TRUE, estimates = TRUE, ci = TRUE)
#significant for veillonella 2, not 18. R2 = 0.24
```

###streptococcus
```{r}
models <- '
temp ~ streptococcus_2_log  + streptococcus_18_log + tc_sex + SES  + months_bf_T18+ PH_cesarean_T2
'
fits <- sem(models, fixed.x=F,data = df_gens)
summary(fits, fit.measures = TRUE, standardized = TRUE, rsquare=TRUE, estimates = TRUE, ci = TRUE)

```
###clostridium
```{r}
modelc <- '
temp ~ clostridium_2_log + clostridium_18_log + tc_sex + SES  + months_bf_T18+ PH_cesarean_T2
'
fitc <- sem(modelc, fixed.x=F,data = df_gens)
summary(fitc, fit.measures = TRUE, standardized = TRUE, rsquare=TRUE, estimates = TRUE, ci = TRUE)#ns

```

###anaerotruncus
```{r}
modelan <- '
temp ~  anaerotruncus_2_log + anaerotruncus_18_log + tc_sex + SES + months_bf_T18+ PH_cesarean_T2 
'
fitan <- sem(modelan, fixed.x=F,data = df_gens)
summary(fitan, fit.measures = TRUE, standardized = TRUE, rsquare=TRUE, estimates = TRUE, ci = TRUE)#ns

```



###ruthenibacterium
```{r}
modelr <- '
temp ~  ruthenibacterium_2_log + ruthenibacterium_18_log + tc_sex + SES + months_bf_T18+ PH_cesarean_T2
'
fitr <- sem(modelr, fixed.x=F,data = df_gens)
summary(fitr, fit.measures = TRUE, standardized = TRUE, rsquare=TRUE, estimates = TRUE, ci = TRUE)#ns

```


#beta diversity analyses
##classical multidimensional scaling

###2 weeks
```{r}
sample_names <- colnames(genus_features2wk)

dist_matrix <- vegdist(genus_features2wk, method = "bray")

# Perform PCoA
pcoa_result <- cmdscale(dist_matrix)

colnames(pcoa_result) <- c("PCoA1", "PCoA2")

genus_beta <- cbind(genus_features2wk, pcoa_result)

beta_id <- genus_beta[,c("SeqID", "PCoA1", "PCoA2")]

df_gens <- merge(df_gens, beta_id, by = "SeqID")

cor.test(df_gens$temp, df_gens$PCoA1) #no correlation
cor.test(df_gens$temp, df_gens$PCoA2) #no correlation

```

###linear model
```{r}
beta_2wk <- '

temp ~ PCoA1 + PCoA2 + tc_sex + SES + months_bf_T18 + PH_cesarean_T2

'
fitbeta2 <- sem(beta_2wk, fixed.x=F,data = df_gens)
summary(fitbeta2, fit.measures = TRUE, standardized = TRUE, rsquare=TRUE, estimates = TRUE, ci = TRUE)

```

###18 months
```{r}
sample_names <- colnames(genus_18mo)

# Calculate dissimilarity matrix (replace this with your own distance calculation method)
# Here, we use Bray-Curtis dissimilarity as an example
dist_matrix <- vegdist(genus_18mo, method = "bray")

# Perform PCoA
pcoa_result <- cmdscale(dist_matrix)

colnames(pcoa_result) <- c("PCoA118", "PCoA218")

genus_beta <- cbind(genus_18mo, pcoa_result)

beta_id <- genus_beta[,c("SeqID", "PCoA118", "PCoA218")]

df_gens <- merge(df_gens, beta_id, by = "SeqID")

cor.test(df_gens$temp, df_gens$PCoA118) #no correlation
cor.test(df_gens$temp, df_gens$PCoA218) #no correlation

```

####linear model
```{r}
beta_18mo <- '

temp ~ PCoA118 + PCoA218 + tc_sex + SES + months_bf_T18 + PH_cesarean_T2

'
fitbeta18 <- sem(beta_18mo, fixed.x=F,data = df_gens)
summary(fitbeta18, fit.measures = TRUE, standardized = TRUE, rsquare=TRUE, estimates = TRUE, ci = TRUE)

```

###linear model both
```{r}
beta_both <- '

temp ~ PCoA1 + PCoA2 + PCoA118 + PCoA218 + tc_sex + SES + months_bf_T18 + PH_cesarean_T2

'
fitbetaboth <- sem(beta_both, fixed.x=F,data = df_gens)
summary(fitbetaboth, fit.measures = TRUE, standardized = TRUE, rsquare=TRUE, estimates = TRUE, ci = TRUE)

```

#Plots
##shannon index
###2 weeks
```{r}
ggplot(df_gens %>% filter(!is.na(temp)), aes(y = temp, x = shannon_2)) +
geom_point(alpha=1, position = position_jitter(height = .002, width = .1), na.rm = TRUE, color='#5533FF') + geom_smooth(method="lm", color='#FF5733') +
  xlab("Shannon Diversity - 2 weeks") + ylab("Negative Affect - 30 months") +
  scale_color_brewer(palette = "Dark2") + #specify colorblind friendly color palette for color
  theme_minimal() +
  #theme( 
  #  panel.grid.major.x = element_blank(),
  #  panel.grid.minor.x = element_blank(),
  #  panel.grid.major.y = element_blank(),
  #  panel.grid.minor.y = element_blank(),
  #) +
  theme(legend.position = "none")

ggsave("figures/shannon2_temp.jpg", height=4, width=6, dpi=1000)


```
###18 months
```{r}
ggplot(df_gens %>% filter(!is.na(temp)), aes(y = temp, x = shannon_18)) +
geom_point(alpha=1, position = position_jitter(height = .002, width = .1), na.rm = TRUE, color='#5533FF') + geom_smooth(method="lm", color='#FF5733') +
  xlab("Shannon Diversity - 18 months") + ylab("Negative Affect - 30 months") +
  scale_color_brewer(palette = "Dark2") + #specify colorblind friendly color palette for color
  theme_minimal() +
  #theme( 
  #  panel.grid.major.x = element_blank(),
  #  panel.grid.minor.x = element_blank(),
  #  panel.grid.major.y = element_blank(),
  #  panel.grid.minor.y = element_blank(),
  #) +
  theme(legend.position = "none")

ggsave("figures/shannon18_temp.jpg", height=4, width=6, dpi=1000)


```

##chao1 index
```{r}
ggplot(df_gens %>% filter(!is.na(temp)), aes(y = temp, x = chao1_2)) +
geom_point(alpha=1, position = position_jitter(height = .002, width = .1), na.rm = TRUE, color='#5533FF') + geom_smooth(method="lm", color='#FF5733') +
  xlab("Chao1 Diversity - 2 weeks") + ylab("Negative Affect - 30 months") +
  scale_color_brewer(palette = "Dark2") + #specify colorblind friendly color palette for color
  theme_minimal() +
  #theme( 
  #  panel.grid.major.x = element_blank(),
  #  panel.grid.minor.x = element_blank(),
  #  panel.grid.major.y = element_blank(),
  #  panel.grid.minor.y = element_blank(),
  #) +
  theme(legend.position = "none")

ggsave("figures/chao2_temp.jpg", height=4, width=6, dpi=1000)


```



##veillonella 2 weeks
```{r}
ggplot(df_gens %>% filter(!is.na(temp)), aes(y = temp, x = veillonella_2_log)) +
geom_point(alpha=1, position = position_jitter(height = .002, width = .1), na.rm = TRUE, color='#7570B3') + geom_smooth(method="lm", color='#E7298A') +
  xlab("Veillonella Relative Abundance (log transformed) - 2 weeks") + ylab("Negative Affect - 30 months") +
  scale_color_brewer(palette = "Dark2") + #specify colorblind friendly color palette for color
  theme_minimal() +
  theme(legend.position = "none")

ggsave("figures/veillonella2_temp_log.jpg", height=4, width=6, dpi=1000)
```

##veillonella 18 months
```{r}
ggplot(df_gens %>% filter(!is.na(temp)), aes(y = temp, x = veillonella_18_log)) +
geom_point(alpha=1, position = position_jitter(height = .002, width = .1), na.rm = TRUE, color='#7570B3') + geom_smooth(method="lm", color='#E7298A') +
  xlab("Veillonella Relative Abundance (log transformed) - 18 months") + ylab("Negative Affect - 30 months") +
  scale_color_brewer(palette = "Dark2") + #specify colorblind friendly color palette for color
  theme_minimal() +
  theme(legend.position = "none")

ggsave("figures/veillonella18_temp_log.jpg", height=4, width=6, dpi=1000)
```

#bifidobacterium 2 weeks
```{r}
ggplot(df_gens %>% filter(!is.na(temp)), aes(y = temp, x = bifidobacterium_2)) +
geom_point(alpha=1, position = position_jitter(height = .002, width = .1), na.rm = TRUE, color='#7570B3') + geom_smooth(method="lm", color='#E7298A') +
  xlab("Bifidobacterium Relative Abundance - 2 weeks") + ylab("Negative Affect - 30 months") +
  scale_color_brewer(palette = "Dark2") + #specify colorblind friendly color palette for color
  theme_minimal() +
  theme(legend.position = "none")

ggsave("figures/bifido2_temp.jpg", height=4, width=6, dpi=1000)

```

#anaerotruncus 18 months
```{r}
ggplot(df_gens %>% filter(!is.na(temp)), aes(y = temp, x = anaerotruncus_18_log)) +
geom_point(alpha=1, position = position_jitter(height = .002, width = .1), na.rm = TRUE, color='#E1BE6A') + geom_smooth(method="lm", color='#40B0A6') +
  xlab("Anaerotruncus Relative Abundance (log transformed) - 18 months") + ylab("Negative Affect - 30 months") +
  scale_color_brewer(palette = "Dark2") + #specify colorblind friendly color palette for color
  theme_minimal() +
  theme(legend.position = "none")

ggsave("figures/anaero18_temp.jpg", height=4, width=6, dpi=1000)

```

#ruthenibacterium 2 weeks
```{r}
ggplot(df_gens %>% filter(!is.na(temp)), aes(y = temp, x = ruthenibacterium_2_log)) +
geom_point(alpha=1, position = position_jitter(height = .002, width = .1), na.rm = TRUE, color='#FFC20A') + geom_smooth(method="lm", color='#0C7BDC') +
  xlab("Ruthenibacterium Relative Abundance (log transformed) - 2 weeks") + ylab("Negative Affect - 30 months") +
  scale_color_brewer(palette = "Dark2") + #specify colorblind friendly color palette for color
  theme_minimal() +
  theme(legend.position = "none")

ggsave("figures/rutheni_2_temp.jpg", height=4, width=6, dpi=1000)

```


#ruthenibacterium 18 months
```{r}
ggplot(df_gens %>% filter(!is.na(temp)), aes(y = temp, x = ruthenibacterium_18_log)) +
geom_point(alpha=1, position = position_jitter(height = .002, width = .1), na.rm = TRUE, color='#FFC20A') + geom_smooth(method="lm", color='#0C7BDC') +
  xlab("Ruthenibacterium Relative Abundance (log transformed) - 18 months") + ylab("Negative Affect - 30 months") +
  scale_color_brewer(palette = "Dark2") + #specify colorblind friendly color palette for color
  theme_minimal() +
  theme(legend.position = "none")

ggsave("figures/rutheni_18_temp.jpg", height=4, width=6, dpi=1000)

```

